---
author: "Ben Bolker and Marta Wayne"
title: "Nearly neutral (?) host-parasite dynamics"
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
---

A first pass at a definition and implementation of a "nearly neutral" (maybe - this is the term we started with but it may not actually be applicable) model of the evolution of a parasite community within a single host population.

Start with a host population of size $N$ and subpopulations $\{n_i\}$ of different pathogen strains (the length of the vector will be dynamic ...).  Each strain has an associated $R_{0,i}$ (the vector *could* be defined in sorted order - I don't know if this would make either the mathematical description or the computational organization more convenient). For practical purposes, we will use the per-contact probability $\beta$ instead, where $R_0 = \beta N/\gamma$ and $\gamma$ is the recovery rate (i.e., probably per time step), set equal for all parasites for the time being (although later we can imagine allowing this to mutate as well). Consider a discrete-time SIS model with mutation in the transmission probability.

1. The first update step is infection: uninfected individuals ($S=N-\sum n_i$) *remain uninfected* with probability
$$
\prod_i (1-\beta_i)^{n_i}
$$
(i.e., a Reed-Frost update). Those individuals that do become infected are subdivided with a per-strain probability of *approximately*
$$
p_i = \frac{n_i R_{0,i}}{\sum_j n_j R_{0,j}} \qquad ,
$$
i.e. a weighted fraction of prevalence. (**Side question #1**: what is the proper probability of being infected first by an individual of type $i$ in this case? ...) If we use the approximation, we could make it stochastic and discrete by drawing a multinomial probability ... or, even more approximately, draw Poisson deviates with mean ($S/N n_i R_{0,i}$) for each infection type ...
2. In the second update step, all *previously* infected individuals recover with probability $\gamma$ (might need to think carefully about the consequences of this particular update rule [infect first, then recover] to see if there are adjustments that need to be made to the relationship between $\beta$, $\gamma$, and $R_0$ ....).
3. Consider mutation of new infecteds. (Steps 2 and 3 can occur in arbitrary order.) Mutation occurs with *per capita* probability $\mu$ (i.e., per new infection). If mutation occurs we pick a new $R_0$ (equivalently, a new $\beta$) with one of the following rules:

- *shift*: $\log R_0 \to \log R_0 + \epsilon$
- *fixed/"house of cards"*: $\log R_0 \to \epsilon$
- *Ornstein-Uhlenbeck*: $\log R_0 \to \log R_0 + \alpha (\widehat{\log R_0} - \log R_0) + \epsilon$ (**note**: if properly parameterized, OU becomes "shift" in the limit as $\alpha \to 0$ and "fixed" as $\alpha \to 1$ ...)

$\epsilon$ is drawn from a Normal distribution with mean $a$ ($<0$) and standard deviation $b$. (We may have to be careful about scaling: if we have discrete time scales then $\beta$ may have to move/be defined on the log-hazard or logit scale instead.)
4. Initial conditions: somewhat arbitrary, but possibly a single strain with $R_{0,1}=2$ and $n_1=0.5$ (which is the equilibrium condition for a single strain).

**Side question #2: what is the effective population size for this type of infection system?** Logically, it seems that $N_e$ should be 1 when $I=1$ (because only one individual is contributing to the population) as well as when $I=N-1$ (because there is only one susceptible left), and presumably maximum ($=N/2???$) when $I=N/2$ ...

## Preliminary results

```{r pkgs,echo=FALSE,message=FALSE}
library(ggplot2); theme_set(theme_bw())
library(viridis)  ## pretty continuous colour scheme
library(GGally)
source("ggpairsfuns.R")
library(tidyr)
library(plyr)
library(dplyr)
library(knitr)
opts_chunk$set(echo=FALSE)
```

```{r melt}
L <- load("nn_runs_1.rda")
all_runs <- lme4:::namedList(res_1000,res_100,res_33)
gamma <- 1/5
all_m <- ldply(all_runs,identity) %>%
    mutate(
        N=as.numeric(gsub("res_","",.id)),
        Iprop=I/N,
        time=time/1000,
        beta=plogis(mean_lbeta)) %>%
    gather(key=var,value=val,-c(time,.id,N)) %>%
    filter(var %in% c("Iprop","beta","mean_lbeta","sd_lbeta"))
```


```{r plot1,fig.width=8,fig.height=8}
g0 <- ggplot(all_m,aes(time,val,colour=factor(N)))+
    geom_line()+facet_wrap(~var,scale="free",nrow=2)+
    scale_colour_brewer(palette="Set1",name="pop size") +
    labs(y="",x="time (x1000 steps)")
g0 %+% subset(all_m,time>100)
```

**Notes**

- parameters:
    - $\gamma$ (recovery period) = 1/5
	- $\mu$ (per-infection mutation prob) = 0.01
	- "shift"-type mutation with a logit-Normal shift probability (see below) with mean=-1, standard deviation=0.5
	- 1 million time steps, reported every 1000 steps
	- initial $R_0$=2 (plot below omits first 100,000 steps, so this isn't visible), except for $N=33$ (boosted to 4 to prevent extinction)
- I'm dealing with $\beta$ (*per capita* infection probabilities) on the logit scale, which makes it much easier to allow mutation on an unconstrained scale (i.e. not worrying about probabilities <0 or >1). Back-transforming summaries of the logit-beta distribution, e.g. going from mean(logit-beta) to mean(beta), is therefore not entirely accurate; we could deal with this in a variety of ways (computing and storing desired summaries on the appropriate scale during the simulation run, or using the delta method). This also means the amount of mutation (on the $\beta$/probability scale) varies according to the base $\beta$; as $\beta$ approaches 0 or 1, the scope for mutation is smaller.
- first panel shows $\beta$; $\beta$ values this close to 1 give rise to *extremely* high $R_0$ values (since $R_0 = \beta N/\gamma$, $1/\gamma=5$ in this example). "Unrealistic" in some sense ... would the results be different if we used an SIRS model (and left individuals in a Resistant state for some period of time)? In any case, it does make sense that $\beta$ is smaller for $N=100$ than $N=1000$ (more drift).
- Not sure why infection proportion is so similar for different $\beta$; would expect that infection proportion would go down? Also, would expect much higher $I^*/N$ ($*$ denotes equilibrium): expected is $1-1/R_0$ ... Is this another effect of variability?  (**New hypothesis**: since we're in discrete time, even if we had 100% infection, a fraction $1/\gamma$ of individuals would be recovering at each time step, so we'd have about a fraction $1-1/\gamma$ infected - $\approx 0.8$ as observed in this case - but why is the infected proportion slightly *higher* than 80%? Is there something funky about the sequence of events/census time, are mutations being counted differently, etc.?)
- standard deviation of logistic-beta is higher in larger population (not just sure if that's what I expected but ...). Not sure if this is a reliable summary (i.e., how is distribution skewed etc.?) Could save more summary statistics about the shape of the distribution, or save a snapshot of the final distribution - saving the entire distribution for every time step could be pretty unwieldy.
- Would be nice to speed up the code (takes about 2-4 minutes per run; increases with pop size, but sublinearly)

These are the distributions of $\beta$ and $\log_{10}(R_0)$ (the latter
certainly needs to be taken with a grain of salt ...)
```{r bdist,fig.width=8,fig.height=4}
avg_vals <- all_m %>% filter(time>750) %>%
    group_by(N,var) %>%
    filter(var %in% c("mean_lbeta","sd_lbeta")) %>%
    summarise(val=mean(val)) %>%
    spread(var,val)
## would be more elegant to construct the logistic-Normal density
## directly ...
gamma <- 1/5
set.seed(101)
randvals <- ddply(avg_vals,"N",
      function(x) with(x,data.frame(beta=plogis(rnorm(1000,mean_lbeta,sd_lbeta))))) %>%
    mutate(`log[10](R[0])`=log10(beta*N/gamma)) %>%
    gather(var,val,-N)
ggplot(randvals,aes(x=val,fill=factor(N)))+
    facet_wrap(~var,scale="free",labeller=label_parsed)+
    geom_density(alpha=0.5)+
    scale_colour_brewer(palette="Set1",name="pop size")
```

## Latin hypercube runs

Run simulations across a broad range of parameters:

```{r LHS,results="as.is"}
lhs_df <- readRDS("lhs_df.rds")
lhs_df %>% gather(var,val,-run) %>%
    group_by(var) %>%
    summarise(min=min(val),max=max(val)) %>% kable(digits=3)
```

For each run, compute the mean infection proportion,
mean(mean logit-beta), and mean(std dev logit-beta) for
the last 10% (last 100,000 steps out of a million).

Haven't spent enough time to figure out what's going on here yet ...

```{r load_runs3,echo=FALSE,cache=TRUE}
fn <- "nn_runs_3.rda"
load(fn)
## get mean values from last 10%
rc1 <- ldply(resList,
             function(x) {
    x %>% filter(time>(0.9*max(time))) %>%
        summarise(
            I=mean(I),
            mean_lbeta=mean(mean_lbeta),
            sd_lbeta=mean(sd_lbeta))
})
## combine with parameter values
rc2 <- cbind(lhs_df[1:nrow(rc1),],rc1)
## scale I to Iprop, add id variable
rc3 <- rc2 %>% mutate(Iprop=I/N,run=1:nrow(rc2)) %>% select(-I)
## organize data for univariate plots (predictor vs response variables)
rc4A <- rc3 %>% select(run,mu,mut_mean,mut_sd,N,gamma) %>%
    gather(var1,val1,-run)
rc4B <- rc3 %>% select(-mu,-mut_mean,-mut_sd,-N,-gamma) %>%
    gather(var2,val2,-run)
lhs_df <- data.frame(run=1:nrow(lhs_df),lhs_df)
rc5 <- full_join(rc4A,rc4B,by="run")
rc6 <- rc5 %>% full_join(lhs_df,by="run") %>% na.omit()
rc3B <- rc3 %>% mutate(hi_mean=mean_lbeta>0)
```

```{r univ_plot,fig.width=10,fig.height=5}
ggplot(rc6,aes(val1,val2))+
    geom_point(aes(colour=mut_mean))+
    facet_grid(var2~var1,scale="free")+    
    scale_color_viridis()
```

```{r pairs_plots,cache=TRUE}
ggpfun <- function(var="mean_lbeta",colscale=scale_color_viridis(),
                   cols=1:5,
                   data=rc3) {
        ggp1 <- ggpairs(data,
                        mapping = ggplot2::aes_string(colour = var),
                        columns=cols,
                        lower = list(continuous = wrap("points",size=1)), ## alpha = 0.3,size=0.5)),
                        diag = list(continuous = "blankDiag"),
                        upper = list(continuous = "blank"))
        return(trim_gg(tweak_colours_gg(ggp1,colscale)))
    }
gList <- lapply(c("mean_lbeta","Iprop","sd_lbeta"),ggpfun)
```

Mean logit-beta:

```{r pairs_mean_lbeta,fig.width=6,fig.height=6}
print(gList[[1]],spacingProportion=0)
```

Mean proportion infected:

```{r pairs_Iprop_lbeta,fig.width=8,fig.height=8}
print(gList[[2]],spacingProportion=0)
```

Mean std dev of logit-beta:

```{r pairs_stddev_lbeta,fig.width=8,fig.height=8}
print(gList[[3]],spacingProportion=0)
```

Compare high-transmission-equilibrium (red) vs low-transmission equilibrium (blue). Add proportion infected to the plot ...

```{r bimodal,fig.width=8,fig.height=8,warning=FALSE}
print(ggpfun(var="hi_mean",colscale=scale_color_manual(values=c("blue","red")),
             cols=c(1:5,8),data=rc3B),spacingProportion=0)
```

## Variable-$\gamma$ models

Preliminary results for allowing infectious period to mutate: everybody gets infected, all the time! Infectious period ($1/\gamma$) just keeps increasing ... (becomes very, very long).

I've also run the latin hypercube runs for the variable-gamma case, but I'm not sure I want to worry about the results until I've understood what's going on in this case. Related to the details of order/what the effective $R_0$ etc. is in this particular discret-time model?

```{r read_Gruns}
L <- load("nn_runs_4.rda")
Nvals <- c(1000,100,33)
names(resG) <- Nvals
all_m <- ldply(resG,identity) %>%
    mutate(
        N=as.numeric(.id),
        Iprop=I/N,
        time=time/1000,
        ## FIXME: will need to use special $linkinv for next runs
        gamma=exp(mean_lgamma)) %>%
    gather(key=var,value=val,-c(time,.id,N)) %>%
    filter(var %in% c("Iprop","gamma","mean_lgamma","sd_lgamma"))
```


```{r plot1G,fig.width=8,fig.height=8}
g0 <- ggplot(all_m,aes(time,val,colour=factor(N)))+
    geom_line()+facet_wrap(~var,scale="free",nrow=2)+
    scale_colour_brewer(palette="Set1",name="pop size") +
    labs(y="",x="time (x1000 steps)")
g0 %+% subset(all_m,time>100)
```

## To do

- figure out bimodality in results for equilibrium transmission
    - seems to have something to do with standard deviation of mutation; low values
	- redo plots as binned average plots (red-blue gradient)
	- redo sims with continuously varying $\gamma$
- understand $\gamma$ results
    - $\gamma$ evolution with logit link, maximum value
	- step through one iteration; asymmetry between $\beta$ and $\gamma$?
- continuous-time equivalent?
- keep track of extinctions?
- SIRS?
- keep track of lineage age?
- output final distribution?
- change the mutational model (e.g. Ornstein-Uhlenbeck, $0 \le \alpha \le 1$)
- make the code faster
